Pattern matching for\\ $(231,213)$-avoiding permutations

### 2015-02-02 article.tex
modification : proposition sur stripe
Section 3

An \emph{ascent} of a permutation $\pi \in S_n$ is any position 
$1 \leq i < n$ where the following value is bigger than the current one. 
That is, if $\pi = \pi_1\pi_2\,\ldots\,\pi_n$, then
$i$ is an ascent if $\pi_i < \pi_{i+1}$.
For example, the permutation 
$345216$7 has ascents (at positions) $1,2,5,6$.
Similarly, a \emph{descent} is a position 
$1 \leq i < n$ with $\pi_i > \pi_{i+1}$, 
so every $i$ with $1 \leq i < n$ either is an ascent or is a descent of 
$\pi$.
Let $\ustep$ and $\dstep$ denote an ascend and a descend, respectively.
The \emph{stripe} $s_\pi$ of a permutation $\pi \in S_n$ is the word
$\stripe{\pi}{1} \stripe{\pi}{2} \ldots \stripe{\pi}{n-1} \in \{\ustep,\dstep\}^{n-1}$ 
defined by 
$ \stripe{\pi}{i}= \ustep$ if $i$ an ascent in $\pi$ and
$\stripe{\pi}{i} = \dstep$ if $i$ a descent in $\pi$.
For example the stripe of the permutation 
$\pi = 12\,11\,10\,1\,2\,3,4\,9\,8\,5\,6\,7$ 
is 
$\stripew{\pi} = \dstep\dstep\dstep\ustep\ustep\ustep\ustep\dstep\dstep\ustep\ustep$.
The stripe $s\stripew{\sigma}$ is a \emph{pattern} of the strip $\stripew{\pi}$
(or $s\stripew{\sigma}$ \emph{occurs} of the strip $\stripew{\pi}$) if $\stripew{\sigma}$ occurs in
$\stripew{\pi}$ as a subsequence.

% For two permutations $\
% The stripe $s\pi$ of $\pi \in S-n$ is a motif pmotif \in \Perm{m}$ is a motif
% of a stripe of $\ptext \in \Perm{n}$ if and only if there exist an increasing function $\phi : [|\pmotif|] \rightarrow [|\pmotif|]$
% such that : \\
% $\forall i \in [|\pmotif|]$, $\mathcal{S}(\pmotif,i) = \mathcal{S}(\ptext,\phi(i))$\\
% If a stripe $S_2$ is a motif of a stripe $S_1$,
% it means that there exists
% a subsequence of $S_1$ such as
% this sequence is equal to $S_2$.

% For example, the stripe $(\dstep_1,\dstep_2,\dstep_3,\dstep_4,\ustep_1)$ is a motif of  the stripe \\
% $(\dstep_1,\dstep_2,\dstep_3,\ustep_1,\ustep_2,\ustep_3,\ustep_4,\dstep_4,\dstep_5,\ustep_5,\ustep_6)$ ,
% because we can match \\
% $(\dstep_1,\dstep_2,\dstep_3,\dstep_4,\ustep_1)$ to $(\dstep_1,\dstep_2,\dstep_3,\dstep_4,\ustep_5,\ustep_6)$.
% \begin{figure}[h!]
% 	\begin{displaymath}
% 		\xymatrix@=0.3cm{
% 			\dstep_1 \ar[dd] & \dstep_2 \ar[dd] & \dstep_3 \ar[dd] & \dstep_4\ar[ddrrrr]  & \ustep_1 \ar[ddrrrrr]\\
% 			\\
% 			\dstep_1 & \dstep_2 & \dstep_3 & \ustep_1 & \ustep_2 & \ustep_3 & \ustep_4 & \dstep_4 & \dstep_5 & \ustep_5 & \ustep_6 \\
% 		}
% 	\end{displaymath}     
% \end{figure}		
% But the stripe $(\dstep,\ustep,\dstep,\dstep,\dstep)$ is not a motif of the stripe\\ $(\dstep_1,\dstep_2,\dstep_3,\ustep_1,\ustep_2,\ustep_3,\ustep_4,\dstep_4,\dstep_5,\ustep_5,\ustep_6,\ustep_7)$.\\

% A \textbf{bivincular motif} generalize
% a motif. 
% A bivincular motif allows 
% to force element to be adjacent in position
% or/and in value.
% A bivincular motif
% is a motif with two set $\bmotif$,
% the first set contains the first index 
% of two consecutive indexes
% and the second set contains the first indexes
% of the two consecutive elements.
% Formally, Let $\bmotif$ 
% be a bivincular motif 
% and $\ptext$ a permutation.
% if $i \in \x$,
% and $i$ is matched with $j$ in $\ptext$
% then $i+1$ must be matched with $j+1$,
% with the special case of $0$ and $|\pmotif|$ :
% if $0 \in \x$ then the first element of $\pmotif$
% must be matched with the first element of $\ptext$,
% and if $|\pmotif| \in \x$ then the last element of $\pmotif$
% must be matched with the last element of $\ptext$.
% if $i \in \y$,
% and $i$ is matched with $j$ in $\ptext$
% then $i+1$ must be matched with $l$
% such as $\ptext(l)=\ptext(j)+1$,
% with the special case of $0$ and $|\pmotif|$ :
% if $0 \in \y$ then the minimal element of $\pmotif$
% must be matched with the minimal element of $\ptext$,
% and if $|\pmotif| \in \y$ then the maximal element of $\pmotif$
% must be matched with the maximal element of $\ptext$.

@@

\begin{proposition}
	\label{proposition:patternMatchingStripe}
	Let $\ptext, \pmotif \in \AV_n(213,231)$.
	$\pmotif$ is a motif of $\ptext$ if and only if the 
	stripe of $\pmotif$ is a motif of the stripe of $\ptext$.
\end{proposition}

		\textbf{Proof} \\
		$\Rightarrow$ Suppose that $\pmotif$ is a motif of $\ptext$
		by subsequence $s$.
		By definition $s$ normalized is equal to
		$\pmotif$ so they have the same stripe.\\ 
		$\Leftarrow$ 
		Let $\ptext$ be a permutation avoiding $(213,231)$, 
		$S_1$ be its stripe, 
		$\pmotif$  be a permutation avoiding $(213,231)$, 
		and $S_2$ its stripe.
		Let $\phi$ be the function
		that match $S_2$ to $S_1$.
		We define $\varphi$ as, 
		$\forall x \in [|\pmotif|-1]$,
		$\varphi(x) = \phi(x)$,
		and $\varphi(|\pmotif|) = \phi(|\pmotif|-1)+1$
		Suppose that $\exists x,y \in [|\pmotif|]$, 
		$\pmotif(x)<\pmotif(y)$ 
		and $\ptext(\varphi(x))>\ptext(\varphi(y))$,
		without lost of generality we can suppose that $x<y$,
		Let see all the case possible that can make this 
		hypothesis possible :
		\begin{itemize}		
			\item if $\pmotif(x)<\pmotif(y)$, 
			$\stripe{\pmotif}{x}=\ustep$ 
			and $\stripe{\pmotif}{y}=\ustep$
			(if x and y are increasing in $\pmotif$),
			then $\stripe{\ptext}{\varphi(x)} =\ustep$ and $\stripe{\ptext}{\varphi(y)}=\ustep$,
			(the match are also increasing).
			As $x<y$ and as $\varphi$ is an increasing function, 
			$\varphi(x) < \varphi(y)$.
			And thus, by \textbf{Lemma \ref{lemma:upup}},
			$\ptext(\varphi(x)) < \ptext(\varphi(y))$ .
			Which contradict our hypothesis,
			so this case is not possible.

@@

			\item if $y = |\pmotif|$ and $x<|\pmotif|-1$,
			if $\stripe{\pmotif}{x}=\ustep$ ie
			$\pmotif(x+1)-\pmotif(x)>0$. Then
			we must have either
			$\pmotif(x)<\pmotif(x+1)<\pmotif(|\pmotif|)$
			or $\pmotif(x)<\pmotif(|\pmotif|)<\pmotif(x+1)$ 
			to avoid $231$ or $213$ in the motif.
			Because we have a match in the stripe
			and what we prove above,
			$\ptext(\varphi(x+1)-\ptext(\varphi(x))>0$.
			But by hypothesis $\ptext(\varphi(|\ptext|))<\ptext(\varphi(x))$,
			so we have either 		
			$\pmotif(|\pmotif|)<\pmotif(x)<\pmotif(x+1)$
			or $\pmotif(|\pmotif|)<\pmotif(x)<\pmotif(x+1)$ 
			which both lead to a $231$ or a $213$ motif
			in the text which is not possible.
			The same idea goes if $\stripe{\pmotif}{x}=\dstep$.

		\end{itemize}
		So our hypothesis is not realisable.So our assertion is true. $\Box$\\

@@

		\section{Pattern Matching With Motif Avoiding (231,213)}

@@

		\section{Pattern Matching With Bivincular Motif Avoiding (231,213)}

@@

	\section{Longest Subsequence Avoiding \\(231,213) for a Permutation}

>>>>

An \emph{ascent} of a permutation $\pi \in S_n$ is any position 
$1 \leq i < n$ where the following value is bigger than the current one. 
That is, if $\pi = \pi_1\pi_2\,\ldots\,\pi_n$, then
$i$ is an ascent if $\pi_i < \pi_{i+1}$.
For example, the permutation 
$345216$7 has ascents (at positions) $1,2,5,6$.
Similarly, a \emph{descent} is a position 
$1 \leq i < n$ with $\pi_i > \pi_{i+1}$, 
so every $i$ with $1 \leq i < n$ either is an ascent or is a descent of 
$\pi$.
Let $\ustep$ and $\dstep$ denote an ascend and a descend, respectively.
The \emph{stripe} $s_\pi$ of a permutation $\pi \in S_n$ is the word
$\stripe{\pi}{1} \stripe{\pi}{2} \ldots \stripe{\pi}{n-1} \in \{\ustep,\dstep\}^{n-1}$ 
defined by 
$ \stripe{\pi}{i}= \ustep$ if $i$ an ascent in $\pi$ and
$\stripe{\pi}{i} = \dstep$ if $i$ a descent in $\pi$.
For example the stripe of the permutation 
$\pi = 12\,11\,10\,1\,2\,3,4\,9\,8\,5\,6\,7$ 
is 
$\stripew{\pi} = \dstep\dstep\dstep\ustep\ustep\ustep\ustep\dstep\dstep\ustep\ustep$.
The stripe $\stripew{\sigma}$ is a \emph{pattern} of the strip $\stripew{\pi}$
(or $\stripew{\sigma}$ \emph{occurs} of the strip $\stripew{\pi}$) if $\stripew{\sigma}$ occurs in
$\stripew{\pi}$ as a subsequence.

@@

\begin{proposition}
	\label{proposition:patternMatchingStripe}
	Let $\sigma \in \AV_k(213,231)$ and $\pi \in \AV_k(213,231)$.
	$\sigma$ is a motif of $\pi$ if and only if 
	$s_\sigma$ is a subsequence of $s_\pi$.
\end{proposition}

\begin{proof}
	The forward direction is essentially obvious.
	For the reverse direction, assume that $s_\sigma$ is a subsequence 
	of $s_\pi$ and
	let $\phi : [k] \to [n]$ be the function describing an occurrence
	of $s_\sigma$ in $s_\pi$ as a subsequence, \emph{i.e.}
	$s_\sigma[i] = s_\pi[\phi(i)]$ for $1 \leq i \leq k$.
	Define $\varphi : [k] \to [n]$ as follows:
	$\varphi(i) = \phi(i)$ for 1 $\leq i \leq k-1$ and
	$\varphi(k) = \phi(k-1) + 1$.
	We show that $\varphi$ describes an occurrence of $\sigma$ in 
	$\pi$.
	Suppose, aiming at a contradiction, that there exist
	$1 \leq i < j \leq k$ such that
	$\sigma[i] < \sigma[j]$ and
	$\pi[\varphi(i)] > \pi[\varphi(j)]$.
	Consider the following cases:

	\begin{enumerate}		
	\item if $\sigma[i] < \sigma [j]$, 
	$s_\sigma[i] = \ustep$ and $s_\pi[j] = \ustep$,
	then $s_\pi[\varphi(i)] = \ustep$ and $s_\pi[\varphi(j)] = \ustep$
	(by construction of $\varphi$).
	Since $i < j$ and $\varphi$ is an increasing function, we have
	$\varphi(i) < \varphi(j)$.
	Therefore, according to Lemma~\ref{lemma:upup},
	$\pi[\varphi(i)] < \pi[\varphi(j)]$.
	This is a contradiction.

@@

	\item if $y = |\pmotif|$ and $x<|\pmotif|-1$,
	if $\stripe{\pmotif}{x}=\ustep$ ie
	$\pmotif(x+1)-\pmotif(x)>0$. Then
	we must have either
	$\pmotif(x)<\pmotif(x+1)<\pmotif(|\pmotif|)$
	or $\pmotif(x)<\pmotif(|\pmotif|)<\pmotif(x+1)$ 
	to avoid $231$ or $213$ in the motif.
	Because we have a match in the stripe
	and what we prove above,
	$\ptext(\varphi(x+1)-\ptext(\varphi(x))>0$.
	But by hypothesis $\ptext(\varphi(|\ptext|))<\ptext(\varphi(x))$,
	so we have either 		
	$\pmotif(|\pmotif|)<\pmotif(x)<\pmotif(x+1)$
	or $\pmotif(|\pmotif|)<\pmotif(x)<\pmotif(x+1)$ 
	which both lead to a $231$ or a $213$ motif
	in the text which is not possible.
	The same idea goes if $\stripe{\pmotif}{x}=\dstep$. 	
	\end{enumerate}
		So our hypothesis is not realisable.So our assertion is true. $\Box$\\		
\end{proof}

		% without lost of generality we can suppose that $x<y$,
% 		Let see all the case possible that can make this 
% 		hypothesis possible :
% 		\begin{itemize}		
% 			\item if $\pmotif(x)<\pmotif(y)$, 
% 			$\stripe{\pmotif}{x}=\ustep$ 
% 			and $\stripe{\pmotif}{y}=\ustep$
% 			(if x and y are increasing in $\pmotif$),
% 			then $\stripe{\ptext}{\varphi(x)} =\ustep$ and $\stripe{\ptext}{\varphi(y)}=\ustep$,
% 			(the match are also increasing).
% 			As $x<y$ and as $\varphi$ is an increasing function, 
% 			$\varphi(x) < \varphi(y)$.
% 			And thus, by \textbf{Lemma \ref{lemma:upup}},
% 			$\ptext(\varphi(x)) < \ptext(\varphi(y))$ .
% 			Which contradict our hypothesis,
% 			so this case is not possible.

% 			\item if $\pmotif(x)<\pmotif(y)$, 
% 			$\stripe{\pmotif}{x}=\ustep$ 
% 			and $\stripe{\pmotif}{y}=\dstep$
% 			then $\stripe{\ptext}{\varphi(x)}=\ustep$ and $\stripe{\ptext}{\varphi(y)}=\dstep$,
% 			As $x<y$ and as $\varphi$ is an increasing function, 
% 			$\varphi(x) < \varphi(y)$.
% 			And thus, by \textbf{Lemma \ref{lemma:updown}},
% 			$\ptext(\varphi(x)) < \ptext(\varphi(y))$ .
% 			Which contradict our hypothesis,
% 			so this case is not possible.

% 			\item if $\pmotif(x)>\pmotif(y)$, 
% 			$\stripe{\pmotif}{x}=\dstep$ 
% 			and $\stripe{\pmotif}{y}=\dstep$
% 			then $\stripe{\ptext}{\varphi(x)}=\dstep$ and $\stripe{\ptext}{\varphi(y)}=\dstep$,
% 			As $x<y$ and as $\varphi$ is an increasing function, 
% 			$\varphi(y) < \varphi(x)$.
% 			And thus, by \textbf{Lemma \ref{lemma:downdown}},
% 			$\ptext(\varphi(y)) > \ptext(\varphi(x))$ .
% 			Which contradict our hypothesis,
% 			so this case is not possible.

% 			\item if $\pmotif(x)>\pmotif(y)$, 
% 			$\stripe{\pmotif}{x}=\dstep$ 
% 			and $\stripe{\pmotif}{y}=\ustep$
% 			then $\stripe{\ptext}{\varphi(x)}=\dstep$ and $\stripe{\ptext}{\varphi(y)}=\ustep$,
% 			As $x<y$ and as $\varphi$ is an increasing function, 
% 			$\varphi(y) < \varphi(x)$.
% 			And thus, by \textbf{Lemma \ref{lemma:updown}},
% 			$\ptext(\varphi(y)) > \ptext(\varphi(x))$ .
% 			Which contradict our hypothesis,
% 			so this case is not possible.$\Box$

% 			\item If $y = |\pmotif|$ and $x=|\pmotif|-1$,
% 			Clearly if $\stripe{\pmotif}{x}=\ustep$,
% 			then $\pmotif(y)-\pmotif(x)>0$ 
% 			and because we have a match in the ribbon
% 			$\ptext(\varphi(y))-\ptext(\varphi(x))>0$.
% 			Same idea goes if $\stripe{\pmotif}{x}=\dstep$.
% 			So this case is not possible.

% 			\item if $y = |\pmotif|$ and $x<|\pmotif|-1$,
% 			if $\stripe{\pmotif}{x}=\ustep$ ie
% 			$\pmotif(x+1)-\pmotif(x)>0$. Then
% 			we must have either
% 			$\pmotif(x)<\pmotif(x+1)<\pmotif(|\pmotif|)$
% 			or $\pmotif(x)<\pmotif(|\pmotif|)<\pmotif(x+1)$ 
% 			to avoid $231$ or $213$ in the motif.
% 			Because we have a match in the stripe
% 			and what we prove above,
% 			$\ptext(\varphi(x+1)-\ptext(\varphi(x))>0$.
% 			But by hypothesis $\ptext(\varphi(|\ptext|))<\ptext(\varphi(x))$,
% 			so we have either 		
% 			$\pmotif(|\pmotif|)<\pmotif(x)<\pmotif(x+1)$
% 			or $\pmotif(|\pmotif|)<\pmotif(x)<\pmotif(x+1)$ 
% 			which both lead to a $231$ or a $213$ motif
% 			in the text which is not possible.
% 			The same idea goes if $\stripe{\pmotif}{x}=\dstep$. 	
% 					\end{itemize}
% 		So our hypothesis is not realisable.So our assertion is true. $\Box$\\		
% \end{proof}

@@

\section{Pattern Matching With Motif Avoiding (231,213)}	
\label{section:sigma only avoid 231 and 213}

@@

\section{Pattern Matching With Bivincular Motif Avoiding (231,213)}	
\label{section:bivincular}

@@

\section{Longest Subsequence Avoiding \\(231,213) for a Permutation}
\label{section:LCS}
